Graphical Tests for Power Comparison of Competing Designs

### 2012-03-31 experiment-two.tex
tweaked results
typos

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=\linewidth]{conf-exp2} 
   \caption{Barcharts of marginal distributions of subjective confidence assessment by design. The distributions display prominent differences.}
   \label{fig:conf-margins}
\end{figure}

Our third response variable is again the subjective assessment of confidence level. On average participants reported a confidence in their results of  3.41 (on a five point scale with 1=lowest and 5=highest), with a standard deviation of 0.149. Modeling confidence levels with a linear mixed effects model with the same structure as before resulted in only one significant effect beyond the intercept: participants report significantly higher confidence for increased number of  points in dotplot lineups. On average an increase of 100 points leads to an increase in confidence of 0.22 (with a standard error of 0.082). This finding is significant with a $p$-value of 0.007. 
Together with the previous results, we can tell that something worth of further investigation happens in the perception of dotplots as sample sizes increase.

Figure \ref{fig:conf-margins} shows barcharts of confidence assessments by design. It is obvious that the distribution of confidence is very different from one design to the next. Boxplots and Dotplots show a similar pattern with an emphasis on high confidence levels, whereas histograms and density lineups have modes at a confidence level of three. 
One reason behind these differences in confidence distributions might be familiarity with the design - neither density charts nor histograms are particularly frequent displays for a general audience.

>>>>

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=\linewidth]{conf-exp2} 
   \caption{Barcharts of marginal distributions of self-reported confidence by design. The distributions display prominent differences.}
   \label{fig:conf-margins}
\end{figure}

Our third response variable is again the subjective assessment of confidence level. On average participants reported a confidence in their results of  3.41 (on a five point scale with 1=lowest and 5=highest), with a standard deviation of 0.149. Modeling confidence levels with a linear mixed effects model with the same structure as before resulted in only one significant effect beyond the intercept: participants report significantly higher confidence for increased number of  points in dotplot lineups. On average an increase of 100 points leads to an increase in confidence of 0.22 (with a standard error of 0.082). This finding is significant with a $p$-value of 0.007. 
Together with the previous results, we can tell that something worthy of further investigation happens in the perception of dotplots as sample sizes increase.

Figure \ref{fig:conf-margins} shows barcharts of confidence assessments by design. It is obvious that the distribution of confidence is very different from one design to the next. Boxplots and dotplots show a similar pattern with an emphasis on high confidence levels, whereas histograms and density lineups have modes at a confidence level of three. 
One reason behind these differences in confidence distributions might be familiarity with the design - neither density charts nor histograms are particularly frequent displays for a general audience.

### 2012-03-31 experiment.tex
typos

Figure \ref{fig:time} shows histogram of the time participants needed to make a decision on each lineup. Correctness of answers is shown by color.  Because of the skewness of the distributions, times were log-transformed. Polar charts take on average more time to answer, but are answered with much lower accuracy. 
The average amount of time spent on a Euclidean lineup is $e^{3.53} = 34.1$ seconds compared to $e^{4.07}=58.5$ seconds for a polar lineup.
This is in stark contrast to accuracy: 76.92\% of the Euclidean lineups shown resulted in the a correct identification of the real data while only 20.31\% of the polar charts were correct.

@@

Figure \ref{fig:conf} shows two barcharts of confidence levels by task, again coloring is used for correctness of answers. Euclidean displays lead to a very bimodal distribution of confidence: participants are either very sure or not sure at all of their answer. Confidence levels in polar charts are distributed much more uniformly.
\begin{figure}[hbtp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=\linewidth]{conf-answer} 
   \caption{Barcharts of self-reported confidence in answering correctly for each lineup. A strongly bimodal distribution is apparent, while confidence levels for polar charts are more uniform. There are only slight (and non significant) difference between confidence levels and correctness of results.}
   \label{fig:conf}
\end{figure}

The perceptual tasks involved in decoding the designs fit into the general framework of the piechart versus barchart discussion. However, we are not dealing with assessing angles, which is known to be a perceptual harder task than assessments of heights or widths in bars, see e.g. \cite{cleveland:1984, robbins:2004, Kosslyn:2006, few:2009}. Instead, the perceptual tasks consist of comparisons along a common axis (in the barcharts) and a common origin (in the polar charts). The difference in designs is therefore based on how well we are able to judge deviations from a horizontal line compared to deviations from a circle. Based on this, the charts with added reference lines provide us with exactly the frame we compare with and should, therefore, be the `better' designs - either in speed or accuracy.

@@

\begin{tabular}{cl}
\phantom{\includegraphics[width=0.03\linewidth]{Polar_Line.pdf}} & \vspace{-0.035in} \multirow{10}{*}{\hspace{-0.5in}\includegraphics[width=0.9\linewidth]{turk4-designs.pdf}} \\
\includegraphics[width=0.045\linewidth]{Polar_Line.pdf} \\
\includegraphics[width=0.045\linewidth]{Polar_NoLine.pdf} \\
\includegraphics[width=0.043\linewidth]{Euclidian_Line.pdf} \\
\includegraphics[width=0.043\linewidth]{Euclidian_NoLine.pdf}\\
\phantom{\includegraphics[width=0.13\linewidth]{Polar_Line.pdf}}\\
\includegraphics[width=0.045\linewidth]{Polar_Line.pdf} \\
\includegraphics[width=0.045\linewidth]{Polar_NoLine.pdf} \\
\includegraphics[width=0.043\linewidth]{Euclidian_Line.pdf} \\
\includegraphics[width=0.043\linewidth]{Euclidian_NoLine.pdf}\\
\phantom{\includegraphics[width=0.13\linewidth]{Polar_Line.pdf}}\\
  \end{tabular} 
  \vspace{0.1in}
   \caption{Power results for two competing designs: polar versus a Euclidean, with and without a reference line; facetting indicates sample size (as percentage of original data). Error bars indicate single Bonferroni adjusted 95\% confidence intervals. The two confidence intervals at the bottom of each panel shows a pairwise power comparison of Euclidean versus polar design according to eqn (\ref{eq:comp}). Differences from zero (vertical grey line) are of interest.}
   \label{fig:treatment}
\end{figure}

>>>>

Figure \ref{fig:time} shows histogram of the time participants needed to make a decision on each lineup. Correctness of answers is shown by color.  Because of the skewness of the distributions, times were log-transformed. Polar charts take on average more time to answer, and are answered with much lower accuracy. 
The average amount of time spent on a Euclidean lineup is $e^{3.53} = 34.1$ seconds compared to $e^{4.07}=58.5$ seconds for a polar lineup.
This is in stark contrast to accuracy: 76.92\% of the Euclidean lineups shown resulted in the a correct identification of the real data while only 20.31\% of the polar charts were correct.

@@

Figure \ref{fig:conf} shows two barcharts of confidence levels by task, again coloring is used for correctness of answers. Euclidean displays lead to a very bimodal distribution of confidence: participants are either very sure or not sure at all of their answer. Confidence levels in polar charts are distributed much more uniformly.
\begin{figure}[hbtp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=\linewidth]{conf-answer} 
   \caption{Barcharts of self-reported confidence in answering correctly for each lineup. A strongly bimodal distribution is apparent, while confidence levels for polar charts are more uniform. Confidence levels are not indicative of accuracy: none of the differences between confidence levels show a significant difference in accuracy.}
   \label{fig:conf}
\end{figure}

The perceptual tasks involved in decoding the designs fit into the general framework of the piechart versus barchart discussion. However, we are not dealing with assessing angles, which is known to be perceptually harder than assessments of heights or widths in bars, see e.g. \cite{cleveland:1984, robbins:2004, Kosslyn:2006, few:2009}. Instead, the perceptual tasks consist of comparisons along a common axis (in the barcharts) and a common origin (in the polar charts). The difference in designs is therefore based on how well we are able to judge deviations from a horizontal line compared to deviations from a circle. Based on this, the charts with added reference lines provide us with exactly the frame we compare with and should, therefore, be the `better' designs - either in speed or accuracy.

@@

\begin{tabular}{cl}
\phantom{\includegraphics[width=0.03\linewidth]{Polar_Line.pdf}} & \vspace{-0.035in} \multirow{10}{*}{\hspace{-0.5in}\includegraphics[width=0.9\linewidth]{turk4-designs.pdf}} \\
\includegraphics[width=0.045\linewidth]{Polar_Line.pdf} \\
\includegraphics[width=0.045\linewidth]{Polar_NoLine.pdf} \\
\includegraphics[width=0.043\linewidth]{Euclidian_Line.pdf} \\
\includegraphics[width=0.043\linewidth]{Euclidian_NoLine.pdf}\\
\phantom{\includegraphics[width=0.13\linewidth]{Polar_Line.pdf}}\\
\includegraphics[width=0.045\linewidth]{Polar_Line.pdf} \\
\includegraphics[width=0.045\linewidth]{Polar_NoLine.pdf} \\
\includegraphics[width=0.043\linewidth]{Euclidian_Line.pdf} \\
\includegraphics[width=0.043\linewidth]{Euclidian_NoLine.pdf}\\
\phantom{\includegraphics[width=0.13\linewidth]{Polar_Line.pdf}}\\
  \end{tabular} 
  \vspace{0.1in}
   \caption{Power results for two competing designs: polar versus  Euclidean, with and without a reference line; facetting indicates sample size (as percentage of original data). Error bars indicate single Bonferroni adjusted 95\% confidence intervals. The two confidence intervals at the bottom of each panel shows a pairwise power comparison of Euclidean versus polar design according to eqn (\ref{eq:comp}). Differences from zero (vertical grey line) are of interest.}
   \label{fig:treatment}
\end{figure}

### 2012-03-31 intro.tex
typos

The power of lineups is used in the work presented here to evaluate the effectiveness of different plot designs. This paper describes two examples where we had real data analysis problems and decisions to make in order to communicate the results. Section \ref{designs} describes the process for comparing designs. Section \ref{setup} and describe the two data analysis problems and the experiments conducted to evaluate the plots designs. Section \ref{results} presents the findings of the experiments, and Section \ref{conclusions} presents conclusions and suggestions for future use.

\section{Comparison of Designs}\label{designs}
We are going to make use of the signal strength gained from multiple viewings of a lineup in order to evaluate competing designs as follows:
\begin{enumerate}
\item{{\bf Create Lineup Data:} create data for a lineup of size $m$ by  creating  $m-1$ permutations of $y$ or, in the case of a simulation study, drawing $m-1$ samples of size $n$ (the number of rows in the data) from the null distribution. Add the original data to the lineup data randomly between 1 and $m$. The {\tt R} package {\tt nullabor} provides a framework for easy creation of lineup data. }
\item{{\bf Create lineups from competing designs:} using the same data, render lineups of all competing designs. }
\item{{\bf Evaluate Lineups:} by presenting the lineups to independent observers. Assess both signal strength and time needed by individuals to come to a decision. Note that each observer should only be exposed to each lineup data once.}
\item{{\bf Evaluate Competing Designs:} differences in signal strength or time to decision are due to differences in the design. In the case that individuals were shown multiple lineups (as part of a bigger study), it is possible to correct outcome measurements for an individual's visual ability. }
\end{enumerate}

>>>>

The power of lineups is used in the work presented here to evaluate the effectiveness of different plot designs. This paper describes two examples where we had real data analysis problems and decisions to make in order to communicate the results. Section \ref{designs} describes the process for comparing designs. Section \ref{setup} describes the two data analysis problems and the experiments conducted to evaluate the plot designs. Section \ref{results} presents the findings of the experiments, and Section \ref{conclusions} presents conclusions and suggestions for future use.

\section{Comparison of Designs}\label{designs}
We are going to make use of the signal strength gained from multiple viewings of a lineup in order to evaluate competing designs as follows:
\begin{enumerate}
\item{{\bf Create Lineup Data:} assuming that at least two variables, $X$ and $Y$ are involved in the design, we create data for a lineup of size $m$ by  creating  $m-1$ permutations of $Y$ or, in the case of a simulation study, drawing $m-1$ samples of size $n$ (the number of rows in the data) from the null distribution. Add the original data to the lineup data randomly between 1 and $m$. The {\tt R} package {\tt nullabor} provides a framework for easy creation of lineup data. }
\item{{\bf Create lineups from competing designs:} using the same data, render lineups of all competing designs. }
\item{{\bf Evaluate Lineups:} by presenting the lineups to independent observers. Assess both signal strength and time needed by individuals to come to a decision. Note that each observer should only be exposed to each lineup data once.}
\item{{\bf Evaluate Competing Designs:} differences in signal strength or time to decision are due to differences in the design. In the case that individuals were shown multiple lineups (as part of a bigger study), it is possible to correct outcome measurements for an individual's visual ability. }
\end{enumerate}

### 2012-03-31 setup.tex
all citet are now cite
typos

Each participant is shown ten lineups and asked to identify the plot that is the most different from the other plots. In the second study the `difference' is further specified as `In which plot is the blue group the furthest to the right?". 
 In both studies, though, the question is placed prominently above each lineup as a reminder of what participants are asked to look for.  Additionally, participants are asked for a reason for their choice. They can select from a list of choices particular to the lineup design, why they chose this particular plot as well as how confident they are on a scale of one to five. Personal information on age group, gender, and education is collected on a voluntary basis.  The amount of time it takes the individual to answer is also recorded.  This  allows us to assess how design choices affect decoding \citet{cleveland:1994} in terms of accuracy and speed.

@@

Statistical tests of mean efficiency by wind direction are not particularly helpful in this situation: the difference in mean efficiency between the  wind direction at which the airport operates the most efficiently and its least efficient direction shows a statistically significant difference with a $p$-value $\le 10^{15}$. %2.2e-16$.  
However, out of the 34 other wind directions,  another 31 show significant differences in efficiency as well. This is much more a property of the large data size rather than practically usable differences. Mere significances also do not allow us an assessment of the underlying pattern.

@@

%%Most generally this experiment can be paralleled to the pie chart versus barchart discussion. The Euclidian coordinate plot is made up of bars *whose?* areas are colored proportionally according to the data. The polar coordinate plot is made up of pie slices which are also colored proportionally according to the data. Viewers of these charts will visually decode the areas of the bars and pie slices and make a judgement based on their decoding. How accurate this judgment is is left to be discussed. According to  \citet[page=40]{kosslyn:2006}, area is usually perceived as a function of what the actual area is. This function is the area raised to an exponent of approximately 0.8 and then multiplied by a constant. However, when bars are parallel, the relative height of these bars is perceived very accurately. From this information, one could infer that the Euclidian coordinate plot may gain efficiency based on the height of the colored bars. 

>>>>

Each participant is shown ten lineups and asked to identify the plot that is the most different from the other plots. In the second study the `difference' is further specified as `In which plot is the blue group the furthest to the right?". 
 In both studies, though, the question is placed prominently above each lineup as a reminder of what participants are asked to look for.  Additionally, participants are asked for a reason for their choice. They can select from a list of choices particular to the lineup design, why they chose this particular plot as well as how confident they are on a scale of one to five. Personal information on age group, gender, and education is collected on a voluntary basis.  The amount of time it takes the individual to answer is also recorded.  This  allows us to assess how design choices affect decoding \cite{cleveland:1994} in terms of accuracy and speed.

@@

Statistical tests of mean efficiency by wind direction are not particularly helpful in this situation: the difference in mean efficiency between the  wind direction at which the airport operates the most efficiently and its least efficient direction shows a statistically significant difference with a $p$-value $\le 10^{-15}$. %2.2e-16$.  
However, out of the 34 other wind directions,  another 31 show significant differences in efficiency as well. This is much more a property of the large data size rather than practically usable differences. Mere significances also do not allow us an assessment of the underlying pattern.

@@

%%Most generally this experiment can be paralleled to the pie chart versus barchart discussion. The Euclidian coordinate plot is made up of bars *whose?* areas are colored proportionally according to the data. The polar coordinate plot is made up of pie slices which are also colored proportionally according to the data. Viewers of these charts will visually decode the areas of the bars and pie slices and make a judgement based on their decoding. How accurate this judgment is is left to be discussed. According to  \cite[page=40]{kosslyn:2006}, area is usually perceived as a function of what the actual area is. This function is the area raised to an exponent of approximately 0.8 and then multiplied by a constant. However, when bars are parallel, the relative height of these bars is perceived very accurately. From this information, one could infer that the Euclidian coordinate plot may gain efficiency based on the height of the colored bars. 
